Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-05-25 doc/invt.tex
adding a few references

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme
  (such as the single-bin and Gaussian schemes commonly used in WL and metadynamics),
  and the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of updating steps with a proper shift,
  as given by the inverse-time prescription.
  %
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
with a general window function,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time
in the single-bin case.
%
Further, we can generalize
the single-bin updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

We can understand the optimality of the inverse-time schedule
as a natural consequence of the stabilization of a runtime average
as the sample size increases\cite{
  marsili2006, barducci2008}.
%
Consider a sample, $\{X^{(1)}, \dots, X^{(N)}\}$,
with a dynamically growing size.
%
The runtime average
%
\begin{equation*}
  {\bar X}^{(N)}
  =
  \frac{ X^{(1)} + \cdots + X^{(N)} } { N }
  ,
\end{equation*}
%
can be written as a recurrence relation regarding
the addition of the last data point,
%
\begin{align}
  {\bar X}^{(N)}
  &=
  {\bar X}^{(N-1)} + \frac{ X^{(N)} - {\bar X}^{(N-1)} } { N }
  .
  \label{eq:av_recur}
\end{align}
%
If we interpret $X^{(N)} - {\bar X}^{(N-1)}$
as the independent correction to the previous average,
${\bar X}^{(N-1)}$, proposed by the latest data point, $X^{(N)}$,
then this correction is absorbed into the average with
a diminishing weight $1/N$ as the sample size, $N$, increases.

>>>>

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of flat-distribution sampling methods
  for free energy calculation including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the error in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme.
  %
  While the asymptotically optimal schedule for
  the single-bin updating scheme (commonly used in WL)
  is given by inverse-time formula,
  that for the Gaussian updating scheme (commonly used in metadynamics)
  is often more complex.
  %and the initial updating magnitude is ideally half of
  %the previous equilibrium value.
  %
  Further,
  we show that the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  These bandpass updating schemes aim at
  partially flattening a few selected histogram modes
  and their optimal schedule
  is also given by the inverse-time prescription.
  %
  Several previous flat-distribution-sampling algorithms
  can be cast as special cases of the bandpass updating schemes. 
  %
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian one
used in the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
Ideally,
the optimal initial updating magnitude
can be set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time
in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter, however,
allow one to selectively flatten only
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also the above origin-shifted inverse-time schedule.
%
Interestingly, several previous FDS algorithms\cite{
  langfeld2012, pellegrini2014,
  neuhaus2006, *neuhaus2007}

%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
%These results may help to better
%plan and optimize FDS simulations.

@@

We can understand the optimality of the inverse-time schedule
as a natural consequence of the stabilization of a runtime average
as the sample size increases\cite{
  marsili2006, barducci2008, robbins1951, pellegrini2014}.
%
Consider a sample, $\{X^{(1)}, \dots, X^{(N)}\}$,
with a dynamically growing size.
%
The runtime average
%
\begin{equation*}
  {\bar X}^{(N)}
  =
  \frac{ X^{(1)} + \cdots + X^{(N)} } { N }
  ,
\end{equation*}
%
can be written as a recurrence relation regarding
the addition of the last data point,
%
\begin{align}
  {\bar X}^{(N)}
  &=
  {\bar X}^{(N-1)} + \frac{ X^{(N)} - {\bar X}^{(N-1)} } { N }
  .
  \label{eq:av_recur}
\end{align}
%
If we interpret $X^{(N)} - {\bar X}^{(N-1)}$
as the independent correction to the previous average,
${\bar X}^{(N-1)}$, proposed by the latest data point, $X^{(N)}$,
then this correction is absorbed into the average with
a diminishing weight $1/N$ as the sample size, $N$, increases.
